---
title: "Covid 19 EDA"
date: "2023-10-23"
---

#About Project
* Exploring Covid event trends in different countries
* Practicing visualization with ggplot
* Transforming data for PowerBI Dashboard




# Table of Contents
* Import Packages
* Data load and EDA
* Visualize different events over time for certain locations
    + China Beijing
    + Australia Capital
    + France Martinique
* Creating a dynamic function to graph 3 events for  any country/ state of choice
* Creating a dynamic function to graph 1 event of choice (confirmed,recovered,deaths) for  any country/ state of choice


# Import Packages
```{r , warning=FALSE ,error=FALSE ,message=FALSE ,results='hide'}

library(data.table)
library(ggplot2)
library(tidyverse)
library(GGally) 

```



# Data load and EDA
```{r}
#Free dataset from link below
#https://www.kaggle.com/datasets/niketchauhan/covid-19-time-series-data
df <- read.table("time-series-19-covid-combined.csv",header=TRUE,sep=",",stringsAsFactors=FALSE)


#Convert to datatable
dt <- as.data.table(df)
dt
```

```{r}
glimpse(dt)
```


Utility Function(s)
```{r}

factorize_str <- function (dt){
  #Purpose:Temporary converts a datatable's character datatypes into factor. This way when summarize function is used, it lists out the counts of the categorical variables
  #input:(datatable)
  #output:(datatable) returns a copy of the datatable but with all the character datatypes as factors
  
  
  #Making a copy to not affect the original datatable
  dt_copy <- copy(dt)

  for (col_name in colnames(dt_copy)){
    if (is.character(dt_copy[[col_name]])){
        set(dt_copy, j = col_name, value = as.factor(dt_copy[[col_name]]))
    }
  }

  #Another way of doing it
  # changeCols <- names(dt_copy)[sapply(dt_copy, is.character)]
  # dt_copy[,(changeCols):= lapply(.SD, as.factor), .SDcols = changeCols]
    
  dt_copy
}

```



```{r}

#split date into 3 sections year,month and day
#This would be used for PowerBI 
dt[, c('year', 'month','day') := tstrsplit(Date, '-', fixed = TRUE,)]

# Print the updated data.table
dt
```



```{r}
summary(factorize_str(dt))
# Lots of empty values for column province.state .... ~50,000/72,000 is approx 70%
# Also some NAs for Recovered column
```



```{r}
cat('There are ',dim(dt)[1] , "columns and ",dim(dt)[2], "rows.")
```


```{r}
#Convert date to date datatypes
dt[,Date:=ymd(Date)]
dt

#Create a new column
dt[ , year_mon:= substring(format(Date, format = "%Y%m%d"),0,6)]
dt

#checking for percentage of nulls for each column
round(colSums(is.na(dt)) * 100/nrow(dt) ,1)

#Checking the number of rows of each column where values are empty strings
colSums(dt=="")
```
```{r} 
# Seems like there is no obvious pattern of why some rows have empty values for 'Province.state' column.
unique(dt[Province.State =="" , .(Country.Region,Province.State) ])

dt[, .N , by=c("Country.Region" , "Province.State")]
```

```{r}
#No recovery data for Canada only. Other countries don't have NA values for Recovery column.
dt[is.na(Recovered)]
summary(factorize_str(dt[is.na(Recovered)]))
```



```{r}
#Date statistics
max(dt$Date)
min(dt$Date)
max(dt$Date)-min(dt$Date) #269 days is ~ 9 months
```

```{r}
#picking some columns we are interested in
country_events_data_dt <- dt[ , .(Country.Region ,Confirmed , Recovered,Deaths)]
country_events_data_dt

```




```{r}

# For each country ,see approximate number of cases per month
country_events_data_dt [ , .(avg_case_per_month=as.integer((sum(Confirmed)*30)/269)), by='Country.Region'][order(-avg_case_per_month)]
#Insight:USA has the most. This could be because USA did lots of testing compared to other countries.

```


```{r}
#See cases per month for each country
country_events_data_dt [ , .(avg_case_per_day=as.integer(sum(Confirmed)/269)), by='Country.Region'][order(-avg_case_per_day)]
```





# Visualize different events over time for certain locations
we will look at some places in China, Australia and France


## Analyzing Beijing, China
```{r}

#Filter for the region of interest
china_beijing_dt <- dt[Country.Region =='China' & Province.State=='Beijing']


summary(factorize_str(china_beijing_dt))


# Getting monthly data for china beijing
china_beijing_agg_dt <- china_beijing_dt [, lapply(.SD, sum), .SDcols=c('Confirmed','Recovered','Deaths') ,by='year_mon']
china_beijing_agg_dt


china_beijing_agg_dt[ ,year_mon:=ym(year_mon)]
china_beijing_agg_dt


```

```{r}

# Seeing distributions and correlations
num.cols <- sapply(china_beijing_agg_dt, is.numeric)
ggpairs( china_beijing_agg_dt[,..num.cols] )
# The correlation makes sense. If there are more cases then there are more opportunities for recovery and death.


```

```{r}

#Seeing monthly cases visually
ggplot(data = china_beijing_agg_dt, mapping = aes(x = year_mon , y=Confirmed )) +
  geom_point() +
  geom_smooth(se = FALSE) +
  ggtitle("China Beijing Year 2020 Monthly Confirmed Cases") +
  labs(y = "Confirmed Cases", x = "month") +
  theme(plot.title = element_text(hjust = 0.5))

```


```{r}
#Visualize Confirmed , Recovery and Death event trends on the same graph
colors <- c("Confirmed" = "blue", "Recovered" = "green" , "Deaths"="red" )
ggplot() + geom_line(aes(x=china_beijing_agg_dt$year_mon,y=china_beijing_agg_dt$Confirmed,color='Confirmed')) + 
           geom_line(aes(x=china_beijing_agg_dt$year_mon,y=china_beijing_agg_dt$Recovered,color='Recovered')) + 
           geom_line(aes(x=china_beijing_agg_dt$year_mon,y=china_beijing_agg_dt$Deaths,color='Deaths')) +
            labs(x = "Month",
                 y = "Events",
                 color = "Legend") + 
           ggtitle("China Beijing Year 2020 Monthly Trio Events") +
           theme(plot.title = element_text(hjust = 0.5)) +
           scale_color_manual(values = colors)

#insight: Very few death. Lots of people recover.
```




## Analysis for Australia Capital
```{r}
Australia_capital_dt <- dt[Country.Region =='Australia' & Province.State =='Australian Capital Territory']
Australia_capital_dt

summary(factorize_str(Australia_capital_dt))
```

```{r}
Australia_capital_agg_dt <- Australia_capital_dt [, lapply(.SD, sum), .SDcols=c('Confirmed','Recovered','Deaths') ,by='year_mon']
Australia_capital_agg_dt

Australia_capital_agg_dt[ ,year_mon:=ym(year_mon)]
Australia_capital_agg_dt
```

```{r}
colors <- c("Confirmed" = "blue", "Recovered" = "green" , "Deaths"="red" )
ggplot() + geom_line(aes(x=Australia_capital_agg_dt$year_mon,y=Australia_capital_agg_dt$Confirmed,color='Confirmed')) + 
           geom_line(aes(x=Australia_capital_agg_dt$year_mon,y=Australia_capital_agg_dt$Recovered,color='Recovered')) + 
           geom_line(aes(x=Australia_capital_agg_dt$year_mon,y=Australia_capital_agg_dt$Deaths,color='Deaths')) +
            labs(x = "Month",
                 y = "Events",
                 color = "Legend") + 
           ggtitle("Australia Year 2020 Monthly Trio Events") +
           theme(plot.title = element_text(hjust = 0.5)) +
           scale_color_manual(values = colors)

#insight: Very few death. Lots of people recover.
```




## Analysis for France Martinique
```{r}

France_martinique_dt <- dt[Country.Region =='France' & Province.State =='Martinique']
France_martinique_dt

summary(factorize_str(France_martinique_dt))
```

```{r}
France_martinique_agg_dt <- France_martinique_dt [, lapply(.SD, sum), .SDcols=c('Confirmed','Recovered','Deaths') ,by='year_mon']
France_martinique_agg_dt 

France_martinique_agg_dt[ ,year_mon:=ym(year_mon)]
France_martinique_agg_dt
```

```{r}
colors <- c("Confirmed" = "blue", "Recovered" = "green" , "Deaths"="red" )
ggplot() + geom_line(aes(x=France_martinique_agg_dt$year_mon,y=France_martinique_agg_dt$Confirmed,color='Confirmed')) + 
           geom_line(aes(x=France_martinique_agg_dt$year_mon,y=France_martinique_agg_dt$Recovered,color='Recovered')) + 
           geom_line(aes(x=France_martinique_agg_dt$year_mon,y=France_martinique_agg_dt$Deaths,color='Deaths')) +
            labs(x = "Month",
                 y = "Events",
                 color = "Legend") + 
           ggtitle("France Year 2020 Monthly Trio Events") +
           theme(plot.title = element_text(hjust = 0.5)) +
           scale_color_manual(values = colors)

#insight: Something weird is happening because cases are high but both recovery and death are low.
```



# Creating a dynamic function to graph 3 events for  any country/ state of choice
```{r}
display_3_event_line_graphs <- function (dt,country,province){
  #Purpose: Display three events for a certain country/province over time
  #inputs:(dt:datatable)
         #(country:character)
         #(province:character)
  #output:a graph
  
  datatable <- copy(dt)
  datatable <- datatable[Country.Region ==country & Province.State ==province]

  datatable_agg_dt <- datatable [, lapply(.SD, sum), .SDcols=c('Confirmed','Recovered','Deaths') ,by='year_mon']
  datatable_agg_dt[ ,year_mon:=ym(year_mon)]

  
  colors <- c("Confirmed" = "blue", "Recovered" = "green" , "Deaths"="red" )
  ggplot() + geom_line(aes(x=datatable_agg_dt$year_mon,y=datatable_agg_dt$Confirmed,color='Confirmed')) + 
             geom_line(aes(x=datatable_agg_dt$year_mon,y=datatable_agg_dt$Recovered,color='Recovered')) + 
             geom_line(aes(x=datatable_agg_dt$year_mon,y=datatable_agg_dt$Deaths,color='Deaths')) +
              labs(x = "Month",
                   y = "Events",
                   color = "Legend") + 
             ggtitle(paste(country," Year 2020 Monthly Trio Events")) +
             theme(plot.title = element_text(hjust = 0.5)) +
             scale_color_manual(values = colors)
}

display_3_event_line_graphs(dt,'France','Martinique')
display_3_event_line_graphs(dt,'Australia','Australian Capital Territory')

```

# Creating a dynamic function to graph 1 event of choice (confirmed,recovered,deaths) for  any country/ state of choice
```{r}

display_event_line_graphs <- function (dt,country,province,metric){
  #Purpose: Display an event for a certain country/province over time
  #input:(dt:datatable)
        #(country:character)
        #(province:character)
        #(metric:character)
  #output:a graph
  
  
  #check if the user inputs the right metric or else break the code!
  if (metric %in% c('Confirmed' ,'Deaths', 'Recovered') == F){ 
      stop("wrong metric column name. Try again!")
  }
  
  datatable <- copy(dt)
  datatable <- datatable[Country.Region ==country & Province.State ==province]

  datatable_agg_dt <- datatable [, lapply(.SD, sum), .SDcols=c(metric) ,by='year_mon']
  datatable_agg_dt[ ,year_mon:=ym(year_mon)]

  ggplot(datatable_agg_dt, aes(x = year_mon, y = get(metric))) + 
    geom_line() + 
    labs(x = "Month",
         y = metric,
         color = "Legend") + 
    ggtitle(paste(country, " Year 2020 Monthly ", metric, " Event")) +
    theme(plot.title = element_text(hjust = 0.5))
}



display_event_line_graphs(dt,'France','Martinique','Recovered')
```


